Tsbrowse: a genome browser for Ancestral Recombination Graphs

### 2025-03-07 manuscript/paper.tex

% Its focus is on interactively visualising the
% information structure in ARGs at all scales.

% \texttt{Tsbrowse} provides interactive visual summaries of  the
% fundamental properties of ARGs, i.e. the mutations, nodes, edges and trees they
% encode.

% \texttt{Tsbrowse} operates on ARGs encoded as succinct tree sequences
% \citep{wong2024general}. 
% This the \texttt{tskit} library~\citep{ralph2020efficiently}
% and supported by 
% simulation~\citep{kelleher2016efficient,kelleher2018efficient,
% haller2019tree,baumdicker2022efficient,
% adrion2020community,lauterbur2023expanding,korfmann2023weak,tsambos2023link,tagami2024tstrait}
% inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
% mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general}
% and downstream processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}

@@

\section{Results} \subsection{Architecture} \texttt{Tsbrowse}
is run in a step-wise approach (Figure \ref{fig:Figure_1}), which makes it
convenient to handle large datasets. The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.

\begin{figure} \centering
\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
defined by a collection of tables which broadly map to ARG properties. Exemplar
table names are denoted in dark blue and columns that describe the property in
light blue. In the pre-process step, tables are augmented with additional
information computed for each property (yellow). The output from this
pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
renders the visualisation in a web browser by leveraging tools in the Holoviz
ecosystem. Optional annotations are provided as an input file, allowing users
to overlay contextual information about genes or other sequence features. }
\label{fig:Figure_1} \end{figure}

\subsection{Data model} \label{subsec:Data_Model} The foundation of
\texttt{tsbrowse} is the \texttt{tskit} data model. As \texttt{tsbrowse} is
based on this generalised ARG representation \citep{wong2024general} it can be applied
to ARGs from several widely-used inference methods including \texttt{Relate}
\citep{speidel2019method}, \texttt{ARG-Needle} \citep{zhang2023biobank} and \texttt{SINGER}
\citep{deng2024robust}. In \texttt{tskit}, ARGs are encoded as a
collection of tables, each of which store information about a specific ARG
property. For example, the Node table describes sampled or ancestral sequences
and records their time of birth, and the individual and population they belong
to. The Edge table describes the relationship between pairs of nodes, and
records the node IDs and the genomic locations over which the edge spans. The
Mutation table records mutation events in the history of the sample, including
the change of allelic state, time of event and node in which the event
occurred. The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, while the Mutation table describes
sequence variation contained in the samples. This compact
representation enables the efficient storage and querying of biobank-scale
ARGs. \texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation (see
\hyperref[subsec:User_Interface]{User Interface}).

>>>>

@@

\section{Results}

\begin{figure} \centering
\includegraphics[width=0.95\linewidth]{figures/MainFig1.png}
\caption{\textbf{Overview of tsbrowse architecture. } In tskit, ARGs are
defined by a collection of tables. Exemplar
table names are denoted in dark blue and columns that describe the property in
light blue. In the pre-process step, tables are augmented with additional
information computed for each property (yellow). The output from this
pre-processing step is stored as a \texttt{.tsbrowse} file. Next the serve step
renders the visualisation in a web browser by leveraging tools in the Holoviz
ecosystem. Optional annotations are provided as an input file, allowing users
to overlay contextual information about genes or other sequence features. }
\label{fig:Figure_1} \end{figure}

\subsection{Data model} \label{subsec:Data_Model} 
\texttt{Tsbrowse} uses the ``succinct tree sequence'' encoding of 
ARGS~\citep{wong2024general}. This efficient ARG encoding 
is implemented by the 
\texttt{tskit} library~\citep{ralph2020efficiently}
and supported by most modern ARG
simulation~\citep{kelleher2016efficient,kelleher2018efficient,
haller2019tree,baumdicker2022efficient,
adrion2020community,lauterbur2023expanding,korfmann2023weak,
tsambos2023link,tagami2024tstrait},
inference~\citep{kelleher2019inferring,speidel2019method,wohns2022unified,
mahmoudi2022bayesian,zhan2023towards,zhang2023biobank,deng2025general},
and processing methods~\citep{fan2022genealogical,nowbandegani2023extremely}.
In \texttt{tskit}, ARGs are encoded as a
collection of tables, storing information about a specific ARG
property (Fig.~\ref{fig:Figure_1}).
% For example, the Node table describes sampled or ancestral sequences
% and records their time of birth, and the individual and population they belong
% to. The Edge table describes the relationship between pairs of nodes, and
% records the node IDs and the genomic locations over which the edge spans. The
% Mutation table records mutation events in the history of the sample, including
% the change of allelic state, time of event and node in which the event
% occurred. 
The Node and Edge tables describe the genealogical relationships
between the sampled and ancestral sequences, 
while the Mutation table describes
sequence variation contained in the samples. 
\texttt{Tsbrowse} augments these tables with additional columns to create
a compressed \texttt{.tsbrowse} file, containing all necessary semantic
metadata required for automatic visualisation.
% (see \hyperref[subsec:User_Interface]{User Interface}).

\subsection{Architecture} 
\texttt{Tsbrowse} is a client-server application (Fig.~\ref{fig:Figure_1}).
The server is a command-line interface program written in Python, which
is run on the computer where the ARG we wish to visualise is available.
The client is a web-browser, which may optionally be running 
on a different machine. There are two important advantages to this 
approach over a monolithic single-machine architecture. Firstly, 
the ARG being visualised remains in-situ and does not need to be 
downloaded from the server; summaries of the ARG at various 
scales are sent to the client on demand.
Secondly, this provides the flexibility to use a powerful server
when generating summaries of ARG while using a standard desktop 
machine for visualisation.

\texttt{Tsbrowse} is run in a step-wise approach (Fig.~\ref{fig:Figure_1}), 
which makes it convenient to handle large datasets. 
The initial step (\texttt{tsbrowse
preprocess}) prepares the data for visualisation. It can be performed
separately on a large machine, and the output stored for later use. The second
step (\texttt{tsbrowse serve}), which can be run on a smaller machine, creates
and renders visualisations. This step follows a client-server architecture. On
the server side (the machine on which \texttt{tsbrowse serve} is run), the
Holoviz suite of libraries \citep{Holoviz} is leveraged to create
visualisations based on the pre-processed data. This choice of architecture is
prompted by the need to represent properties of large-scale ARGs in a
meaningful and performant way. For example, a visual representation of millions
of mutations in SARS-Cov-2 ARGs \citep{zhan2023towards} could quickly
lead to performance bottlenecks and overplotting. Datashader
\citep{Datashader}, a component Holoviz package, overcomes this challenge by
rasterising huge datasets (see \hyperref[subsec:User_Interface]{User Interface}
for details, and Supplementary Figure \ref{fig:Supplementary_Figure_4} for a
tsbrowse view of SARS-Cov-2 mutations). A web browser client connects to the
server to render and interact with these visualisations. Heavy data processing
is limited to the server side, keeping the client-side interface lightweight.
For large datasets with complex visualisations, this achieves a smoother user
experience.
